Forecasting biodiversity in breeding birds using best practices

### 2017-12-11 manuscript/reviewer_reply.Rmd

We would like to thank the editor and the two reviewers for the productive feedback on our manuscript. We have addressed the comments related to our choice of data, clarified the results, added more description to our data and methods, and included more language on future directions of this research. More detailed responses to individual comments are described below.

### 2017-12-11 manuscript/reviewer_reply.Rmd

*Line 62: I really like the recommendations / clarity in Box 1, and think these will be accessible to most ecologists*

>Thank you.

*Line 71-73: I’m re-reading this after commenting on data, and am now a little confused. The bird responses are annual snapshots during the breeding season? And I thought the environmental data you used was annual or quarterly, not monthly?* https://www.mbr-pwrc.usgs.gov/bbs/grass/genintro.htm

>Bird data is collected once each year during the breeding season. Climate data is available monthly. We want to use the chunk of the climate data that is mostly likely to predict the the bird response. We used climate values from the 12 months leading up to the bird sampling since this is assumed to influence the habitat and other resources the birds are responding to (We have also clarified this in the methods). Other choices are possible, but rarely explored b/c what is most commonly used is long-term averages. Future work could explore shorter time scale environmental data such as monthly NDVI trends (instead of just means), migration indicators for migratory species, or more localized weather around each sampling site that could be incorporated into an observer model.

@@

*Line 245: In our 2014 forecasting paper, we used MASE over RMSE because of rationale in Hyndmand and Koehler (2006). Using RMSE here is probably ok because you probably don’t have zeros/huge values/etc*

>We agree.

*Line 249: Can you elaborate on what you did with the deviance? By itself mean deviance is just a measure of average fit (I think most of the models are giving you this). Var(deviance) should incorporate precision – but I’ve only seen this extracted from Bayesian models*

>We were unclear here. We had only intended to convey that mean deviance was lowest for predictions that were made with appropriate uncertainty. We did not use Var(deviance). The new sentence should be clearer on this point.

### 2017-12-11 manuscript/reviewer_reply.Rmd

We would like to thank the editor and the two reviewers for the productive feedback on our manuscript. We have addressed the comments related to our choice of data, clarified the results, added more description to our data and methods, and included more language on future directions of this research. More detailed responses to individual comments are included below.

### 2017-12-11 manuscript/reviewer_reply.Rmd

*Line 62: I really like the recommendations / clarity in Box 1, and think these will be accessible to most ecologists*

>Thank you.

*Line 71-73: I’m re-reading this after commenting on data, and am now a little confused. The bird responses are annual snapshots during the breeding season? And I thought the environmental data you used was annual or quarterly, not monthly?* https://www.mbr-pwrc.usgs.gov/bbs/grass/genintro.htm

>Bird data is collected once each year during the breeding season. Climate data is available monthly. We want to use the chunk of the climate data that is mostly likely to predict the bird response. We used climate values from the 12 months leading up to the bird sampling since this is assumed to influence the habitat and other resources the birds are responding to (We have also clarified this in the methods). Other choices are possible, but rarely explored b/c what is most commonly used is long-term averages (e.g., 30 year climate means). Future work could explore shorter time scale environmental data such as monthly NDVI trends (instead of just means), migration indicators for migratory species, or more localized weather around each sampling site that could be incorporated into an observer model.

### 2017-12-11 manuscript/reviewer_reply.Rmd

>We have switched the panels and labeled them. Thank you for noticing this error.

*I will leave the remainder of the comments to the two reviewers.*

## Reviewer 1: Eric Ward
#### Basic reporting

*I found the paper very clear and easy to follow, with a good intro/discussion that tied the current work into previous studies. All data shared - in addition to code. More minor comments in General Comments section.*
#### Experimental design

*This question and forecasting competition was well designed. I had minor comments to improve clarity with models. The analysis is all documented/replicable on the Github repo. More minor comments in General Comments section.*
#### Validity of the findings

@@

>See comment above on using shorter time scale environmental variables.

*Line 124: I’m not as familiar with terrestrial projections, but are the CMIP5 projections relatively unbiased / precise when compared to recent data?*

>The accuracy of CMIP5 models varies by the variable considered. Temperature is represented very well over North America. Precipitation is challenging, especially over mountainous terrain, but the CMIP5 ensemble represents precipitation across North America relatively well ("Evaluation of precipitation and temperature simulation performance of the CMIP3 and CMIP5 historical experiments", Koutroulis et al. 2016).

*Line 138: Maybe also cite any of the interfaces you used, like RStan: http://mc-stan.org/users/citations/*

@@

*Line 269: See my above comments about spatially correlated random effects / GMRF models. One way to confirm these other approaches are overkill would be to make some maps of the estimated random effects, and fit some simple models to estimate variogram parameters – e.g. even gls() in R can do this*

>See our comments about spatial autocorrelation above

*Line 276: Is one of the potential reasons that richness is stationary that presence/absence on the BBS varies less than other measures – maybe abundance, or other measures of diversity are more variable?*

>We agree that this is likely. We make this point on lines 423-429 of the revised manuscript.

@@

*the authors perform their analyses on data that are collected from an incredibly dynamic time period. I wouldn’t expect species to have settled into a richness pattern given the incredibly disrupted nature of the last 32 years. Some strategies to try to handle these issues might be to train and predict at different temporal scales and to both forecast and hindcast the predictions. (e.g. train on 2004-2013 and try to predict 1982-1992 data + 1993-2003 data). If the authors were to make splits based on expected periods of dynamism, they could better tell whether the drivers that are included in the models are accurate predictors.*

>We agree that understanding the conditions under which forecasts are likely to perform well is an important step for ecological forecasting. We have added a sentence to the discussion making this point. That said, most ecological forecasts are intended to be applied to the future, which at least at the moment we would expect to remain "incredibly dynamic". As such the dynamics of the last decade are likely a reasonable representation of the kinds of dynamics we are trying to forecast.

*124-130: I didn’t see where future projections were performed/used in this paper. Why is this here?*

>Thank you for pointing out that we were unclear here.  The methods and purpose for the future predictions are now explained in more detail on lines 136-139 and 502-505 of the revised manuscript.

@@

*Fig. 6 caption: “compoonents”*

>Thank you for pointing out this error. We have fixed it

*466: only other mention of future forecasts… not sure why this is included in this paper.*

> Thank you again for pointing out that we were unclear here. The purpose of the future forecasts is now explained on lines 136-139 and 502-505.

### 2017-12-11 manuscript/reviewer_reply.Rmd

---
output:
  pdf_document: default
geometry: left=2.5cm, right=2.5cm, top=2.25cm, bottom=2.25cm, headheight=12pt, letterpaper
fontsize: 12pt
---

\raggedright

@@

#### Review of Harris et al. Forecasting biodiversity in breeding birds using best practices

*In general, I thought this paper was interesting and well written. These results add to the increasing literature supporting the idea that ecological forecasting in time/space is difficult. I think the recommendations/best practice bullet points described by the authors are essential for anyone interested in forecasting this kind of data. All of my points below can be considered minor issues, which I think will improve the paper’s clarity.*

>Thank you.

#### Introduction:

*Line 41: I think this is generally more common in terrestrial ecology – in fisheries these SDM models with more of a theoretical background don’t seem to be used as much*

### 2017-12-11 manuscript/draft.Rmd
Add new retriever reference

**Richness data.** Bird species richness was obtained from the North American
Breeding Bird Survey (BBS) [@pardieck2017] using the Data Retriever Python
package [@morris2013] and rdataretriever R package [@rdataretriever]. BBS
observations are three-minute point counts made at 50 fixed locations along a
40km route. Here we denote each route as a site and summarize richness as the
total species observed at all 50 locations in each surveyed year. Prior to
summarizing the data was filtered to exclude all nocturnal, cepuscular, and
aquatic species [since these species are not well sampled by BBS methods;
@hurlbert2005], as well as unidentified species, and hybrids. All data from
surveys that did not meet BBS quality criteria were also excluded.

@@

We evaluated each model's forecasts using the data for each year between `r
timeframe$last_train_year + 1` and `r timeframe$end_yr`. We used three metrics
for evaluating performance: 1) root-mean-square error (RMSE) to determine how
far, on average, the models' predictions were from the observed value; 2) the
95% prediction interval coverage to determine how well the models predicted the
range of possible outcomes; and 3) deviance (i.e. negative 2 times the Gaussian
log-likelihood) as an integrative measure of fit incorporating good point
estimates, precision, and coverage. In addition to evaluating forecast
performance in general, we evaluated how performance changed as the time horizon
of forecasting increased by plotting performance metrics against year. Finally,
we decomposed each model's squared error into two components: the squared error
associated with site-level means and the squared error associated with annual
fluctuations in richness within a site. This decomposition describes the extent
to which each model's error depends on consistent differences among sites versus
changes in site-level richness from year to year.

>>>>

**Richness data.** Bird species richness was obtained from the North American
Breeding Bird Survey (BBS) [@pardieck2017] using the Data Retriever Python
package [@morris2013; @senyondo2017] and rdataretriever R package
[@rdataretriever]. BBS observations are three-minute point counts made at 50
fixed locations along a 40km route. Here we denote each route as a site and
summarize richness as the total species observed at all 50 locations in each
surveyed year. Prior to summarizing the data was filtered to exclude all
nocturnal, cepuscular, and aquatic species [since these species are not well
sampled by BBS methods; @hurlbert2005], as well as unidentified species, and
hybrids. All data from surveys that did not meet BBS quality criteria were also
excluded.

@@

We evaluated each model's forecasts using the data for each year between `r
timeframe$last_train_year + 1` and `r timeframe$end_yr`. We used three metrics
for evaluating performance: 1) root-mean-square error (RMSE) to determine how
far, on average, the models' predictions were from the observed value; 2) the
95% prediction interval coverage to determine how well the models predicted the
range of possible outcomes; and 3) deviance (i.e. negative 2 times the Gaussian
log-likelihood) as an integrative measure of fit that incorporates both accuracy
and uncertainty. In addition to evaluating forecast performance in general, we
evaluated how performance changed as the time horizon of forecasting increased
by plotting performance metrics against year. Finally, we decomposed each
model's squared error into two components: the squared error associated with
site-level means and the squared error associated with annual fluctuations in
richness within a site. This decomposition describes the extent to which each
model's error depends on consistent differences among sites versus changes in
site-level richness from year to year.
